---
title: "Assignment 3: Data Exploration"
author: "Vignesh Arunkumar"
date: "Fall 2024"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Canvas.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Load necessary packages (tidyverse, lubridate, here), check your current working directory and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r Setup,  message = FALSE}
#Loading Library
library(tidyverse)
library(lubridate)
library(here)

#Getting working directory
getwd()

#reading csv files
Neonics<-read.csv("~/EDE_Fall2024/Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv", stringsAsFactors = TRUE)
Neonics

Litter<-read.csv("~/EDE_Fall2024/Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv", stringsAsFactors = TRUE)
Litter


```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer:There are a few different reasons. The main one could be to view the mortality of key species needed in agriculture and pollination or pest control and to eventually prevent it by changing insecticides or finding alternatives. 

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer: There are a few different reasons to be interested in studying litter and woody debris that fall to the ground in the forest. In a foresters perspective we may want to monitor the amount of fuel build up that occurs on the bottom of a forest. From there we can assign our own burns depending on the quanity of fuel/debris to prevent extreme fires that have heavy fuel loads to get out of hand. From an ecological research point of view we may want to see how much of this organic matter is on the forest floor which can provide habitat and nutrients for a variety of organisms. 

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1. Sampling only occurs in terrestrial NEON sites with woody vegetation that is greater than 2 meters in height. 
 2.Sites with forested tower airsheds, litter sampling is taken in 4 40mx40m plots and 26 20mx20m plots.
 3. Depending on vegetation trap placement must be randomized or targeted. For example Vegetation with heights greater than 2 m which cover more than 50% of the forest must have randomized trap placement. 

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r Dimensions}
# dimensions
dim(Neonics)

```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest? [Tip: The `sort()` command is useful for listing the values in order of magnitude...]

```{r Sorting Effects Column}

#sorting effects column
sort(summary(Neonics$Effect))

```

> Answer: The most common effects that are studied are Population and Mortality. This may be to monitor population growth overtime to see any sudden declines in species due to the exposure. 


7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: Explore the help on the `summary()` function, in particular the `maxsum` argument...]

```{r 6most common species}
#sorted the column to see the 6 most common species decreasing in magnitude

sort(summary(Neonics$Species.Common.Name, maxsum = 7), decreasing = TRUE)
```

> Answer: Honey Bee, Parasitic Wasp,  Buff Tailed Bumblebee, Carniolan Honey Bee, Bumble Bee, Italian Honeybee. These species all occur in the Hymenoptera order of insects. After running the same code but with maxsum set to 9 to see what comes after, I am sure that these are more important due to their role as pollinators. 

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric? [Tip: Viewing the dataframe may be helpful...]

```{r Class of column}

#finding what class the column is
class(Neonics$Conc.1..Author.)

```

> Answer: Factor, there are other values other then numbers. 


## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r Graph of Neonics publication count by year}
#making graph of Neonics to view publication count by the year 
help("geom_freqpoly")
ggplot(Neonics, aes(x= Neonics$Publication.Year)) +
  geom_freqpoly()

```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r Neonics Graph with different test locations}
#separating data by test location 
ggplot(Neonics, aes(x= Neonics$Publication.Year, colour = Neonics$Test.Location)) +
  geom_freqpoly()

```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer:Lab, Generally they all increase overtime. The most common it looks like is Lab especially in the last decade. however field natural is in higher counts some years. 

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r Bar graph}
#creating bar graph and adjusting x axis labels
ggplot(Neonics, aes(x= Endpoint)) +
  geom_bar() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

> Answer: LOEL, terrestrial, "Lowest-observable-effect-level: lowest dose (concentration)
producing effects that were significantly different (as reported
by authors) from responses of controls (LOEAL/LOEC) (GDIT 722)" and NOEL, terrestrial, "No-observable-effect-level: highest dose (concentration)
producing effects not significantly different from responses of
controls according to author's reported statistical test
(NOEAL/NOEC) (GDIT 721)". 

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r Changing column class}
#finding column class and changing the class to date
class(Litter$collectDate)
Litter$collectDate<-as_date(Litter$collectDate)
class(Litter$collectDate)
unique(Litter$collectDate)
```

13. Using the `unique` function, determine how many different plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r Identifying different plots}
#Comparing unique and summary functions when calling on a column
unique(Litter$plotID)
summary(Litter$plotID)
```

> Answer: 12, Unique gives all the different plots where as summary allows the count of samples taken as well in its function. 

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r Bar graph for counts of Different functional groups}
# Creating bar graph for counts of functional groups
ggplot(Litter, aes(x= functionalGroup)) +
  geom_bar() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r Box plot and Violin Plots}
# Creating Blox plots and violin plot to compare functional groups and drymass
ggplot(Litter, aes(x= functionalGroup , y = dryMass)) +
  geom_boxplot() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

ggplot(Litter, aes(x= functionalGroup , y = dryMass)) +
  geom_violin() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer:Boxplot allows you to see the range outliers and the average as well as quartiles. Violin plot would be better with other forms of data. 

What type(s) of litter tend to have the highest biomass at these sites?

> Answer: Highest Biomass are needles. 
