---
title: "Assignment 10: Data Scraping"
author: "Vignesh Arunkumar"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 
library(tidyverse)
library(rvest)

getwd()
```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2023 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2023>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 
lwsp.url <- 
"https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2023"

lwsp_durham <- read_html(lwsp.url)

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 
water_system_name <- lwsp_durham %>% 
  html_element("td tr:nth-child(1) td:nth-child(2)") %>% 
  html_text(trim = TRUE)

pwsid <- lwsp_durham %>% 
  html_element("td tr:nth-child(1) td:nth-child(5)") %>% 
  html_text(trim = TRUE)

ownership <-  lwsp_durham %>% 
  html_element("tr:nth-child(2) td:nth-child(4)") %>% 
  html_text(trim = TRUE)

max_day_use <- lwsp_durham %>% 
  html_elements(":nth-child(31) td:nth-child(9),
                tr:nth-child(2) :nth-child(9), :nth-child(31) td:nth-child(6),
                :nth-child(31) td:nth-child(3)") %>% 
  html_text(trim = TRUE)


```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... Or, you could scrape month values from the web page...

5. Create a line plot of the maximum daily withdrawals across the months for 2023, making sure, the months are presented in proper sequence.

```{r create.a.dataframe.from.scraped.data}
#4 
months <- c("Jan" ,"May", "Sept", "Feb", "Jun", "Oct", "Mar", "Jul", "Nov",
            "Apr", "Aug", "Dec")
months.num <- c(01, 05, 09, 02, 06, 10, 03, 07, 11, 04, 08, 12)
years <- rep(2023, 12)
lwsp.dataframe <- data.frame("Water System Name" = water_system_name, 
                             "PWSID" = pwsid,
                             "Ownership" = ownership, 
                             "Max Day Use" = as.numeric(max_day_use) ,
                             "Month" = months, 
                             "Date"= make_date(years, months.num))
lwsp.dataframe <- arrange(lwsp.dataframe, Date)
  
#5 
ggplot(lwsp.dataframe, aes(x= Date, y= Max.Day.Use)) +
  geom_point()+
  geom_line()+
labs(x = "Date",y = "Maximum Daily Witdrawals")


```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data, returning a dataframe.
**Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function}
#6. 
scrape.it <- function(ID, Yr){
  url <- 
gsub(" ","",paste("https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=",ID,"&year=",
    Yr))
read.my.url <- read_html(url)
print(url)
#scrapedata
system_name <- read.my.url %>% 
  html_element("td tr:nth-child(1) td:nth-child(2)") %>% 
  html_text(trim = TRUE)

the.id <- read.my.url %>% 
  html_element("td tr:nth-child(1) td:nth-child(5)") %>% 
  html_text(trim = TRUE)

owner <-  read.my.url %>% 
  html_element("tr:nth-child(2) td:nth-child(4)") %>% 
  html_text(trim = TRUE)

maxuseday <- read.my.url %>% 
  html_elements(":nth-child(31) td:nth-child(9) , tr:nth-child(2) :nth-child(9),
                :nth-child(31) td:nth-child(6),
                :nth-child(31) td:nth-child(3)")%>% 
  html_text(trim = TRUE)



#creates dataframe
months <- c("Jan" ,"May", "Sept", "Feb", "Jun", "Oct", "Mar", "Jul", "Nov",
            "Apr", "Aug", "Dec")
months.num <- c(01, 05, 09, 02, 06, 10, 03, 07, 11, 04, 08, 12)
years <- rep(Yr, 12)
df <- data.frame("Water System Name" = system_name, 
                             "PWSID" = the.id,
                             "Ownership" = owner, 
                             "Max Day Use" = as.numeric(maxuseday) ,
                             "Month" = months, 
                             "Date"= make_date(years, months.num))
df <- arrange(df, Date)

print(df)
return(df)

#plot daily use
ggplot(df, aes(x= Date, y= Max.Day.Use)) +
  geom_point()+
  geom_line()+
labs(x = "Date",y = "Maximum Daily Witdrawals")
}

```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7 
durham15 <- scrape.it('03-32-010', '2015')

```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.






```{r fetch.and.plot.Asheville.2015.data}
#8 
ash <- scrape.it('01-11-010', '2015')

#My path is changed in the function for max use day. Which prevents me from
#getting every months max use day value. 

scrape.it.asheville <- function(ID, Yr){
  url <- 
gsub(" ","",paste("https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=",ID,"&year=",
    Yr))
read.my.url <- read_html(url)
print(url)
#scrapedata
system_name <- read.my.url %>% 
  html_element("td tr:nth-child(1) td:nth-child(2)") %>% 
  html_text(trim = TRUE)

the.id <- read.my.url %>% 
  html_element("td tr:nth-child(1) td:nth-child(5)") %>% 
  html_text(trim = TRUE)

owner <-  read.my.url %>% 
  html_element("tr:nth-child(2) td:nth-child(4)") %>% 
  html_text(trim = TRUE)

maxuseday <- read.my.url %>% 
  html_elements(":nth-child(30) td:nth-child(9) , :nth-child(30) td:nth-child(6), :nth-child(30) td:nth-child(3), tr:nth-child(2) :nth-child(6) tr:nth-child(2) :nth-child(3) tr:nth-child(2) td:nth-child(9)")%>% 
  html_text(trim = TRUE)

#creates dataframe
months <- c("Jan" ,"May", "Sept", "Feb", "Jun", "Oct", "Mar", "Jul", "Nov",
            "Apr", "Aug", "Dec")
months.num <- c(01, 05, 09, 02, 06, 10, 03, 07, 11, 04, 08, 12)
years <- rep(Yr, 12)
df.ash <- data.frame("Water System Name" = system_name, 
                             "PWSID" = the.id,
                             "Ownership" = owner, 
                             "Max Day Use" = as.numeric(maxuseday) ,
                             "Month" = months, 
                             "Date"= make_date(years, months.num))
df.ash <- arrange(df.ash, Date)

print(df.ash)
return(df.ash)

#plot daily use
ggplot(df.ash, aes(x= Date, y= Max.Day.Use)) +
  geom_point()+
  geom_line()+
labs(x = "Date",y = "Maximum Daily Witdrawals")
}

asheville15 <- scrape.it.asheville('01-11-010', '2015')
durham15

combined.data<- merge(asheville15, durham15, all.x = TRUE, all.y = TRUE)

ggplot(combined.data, aes(x= Date, y= Max.Day.Use, color = Water.System.Name)) +
  geom_point()+
  geom_line()+
labs(x = "Date",y = "Maximum Daily Witdrawals")
```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2018 thru 2022.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data }
#9 
#this is not working
the_years = seq(2018,2022)
the_site = rep('01-11-010', length(the_years))
ash.max.water <- map2(the_site, the_years, scrape.it.asheville) %>%
  bind_rows()

#this is working with the original function
the_years = seq(2018,2022)
the_site = rep('01-11-010', length(the_years))
ash.max.water <- map2(the_site, the_years, scrape.it) %>%
  bind_rows()

ash.max.water %>%
ggplot(aes(x= Date, y= Max.Day.Use)) +
#geom_line() +
geom_smooth(method='loess',se=FALSE) +
scale_x_date(date_breaks = '1 year', date_labels = '%Y')

```

Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?
> Answer:
>
